Draft 1

### 2020-05-24 Paper.Rmd
New datasets and equations

---
title: "Paper"
author: "Ana Karina"
date: "4/7/2020"
output: html_document
bibliography: references.bib
---

@@

# Risk of infection from touching surfaces contaminated with SARS-CoV-2

## Abstract

## Introduction

@@

@@

Therefore, touching contaminated surfaces offers a risk for exposure. 
Inspite its importance, there is not a lot of information regarding the magnitud of risk associeted with touching contaminated surfaces @BarOn2020.

@@

## Materials and Methods 
### Hazard Identification
This risk assessment is designed to explore the risks associated indirect transmission of SARS-CoV-2.

Air samples negative, but surface samples positive @Ong2020.
Human coronaviruses can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020.

### Exposure Assessment 
Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.

Two expossure scenarios were modeled. 
*Scenario 1. Single touch.*
Frequently touched buttons (Traffic light buttons, train buttons). Supermarket products.

@@

@@

*Conversing viral RNA data to infectious viruses*
There was a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals. Therefore, we had to translate the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data. Althoug not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF). 
To do this we ...

*Viral shedding and infection potential* 
Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020. The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF). In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.

### Dose Response for SARS-CoV
“For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”

The dose-response relationship was obtain from QMRA wiki <http://qmrawiki.org/experiments/sars>. The model is based on animal dose-response relationship. No human dose-response relationship is available for SARS Coronavirus. It is generally accepted to use animal models in QMRA if human data is not available (REF)

@@

### Limitations
Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.

@@

## Supporting Information
## Author Information
## Aknowledgements 
## References

#### More info that could be added

Large droplets (>= 5um) that fall rapidly near the source contaminate proxime surfaces, making indirect human-surface transmission.

Fecal-oral route has been suggested as a potential transmission route for SARS-CoV-2 @Yeo2020.
“Toilet bowl and sink samples were positive, suggesting that viral shedding in stool” @Ong2020.

"“aerosol and fomite transmission of SARS-CoV-2 is plausible, since the virus can remain viable and infectious in aerosols for hours and on surfaces up to days (depending on the inoculum shed).” @VanDoremalen2020

Part of the WHO COVID-19 strategy is to "engaging and educating the public to ensure protective measures are implemented effectively as changes are made to existing policies and restrictions." @WHO2020a

He et al. analized the patterns of viral sheddin of 94 COVID-19 patients. They also suggested substantial presymptomatic transmission @He2020. "we have estimated that viral shedding of patients
with laboratory-confirmed COVID-19 peaked on or before symptom onset, and a substantial proportion of transmission probably occurred before first symptoms in the index case. More inclusive criteria for contact tracing to capture potential transmission events 2 to 3 days before symptom onset should be urgently considered for effective control of the outbreak"

Matrices: saliva, mucus, nasal secretions, lower respiratory tack secretions, urine, blood.

>>>>

---
title: "Draft 1"
author: "Ana Karina"
date: "4/7/2020"
output: html_document
bibliography: references.bib
---

@@

# Risk of infection from touching surfaces contaminated with SARS-CoV-2
## Abstract
## Introduction

extensive environmental contamination has been found in settings with COVID-19 patients @Ong2020 @Ye2020, coronaviruses persist on inanimate surface for extended periods of time @Kampf2020

@@

Surfaces play a significant role in the spread of respiratory viral diseases @Boone2007

Potential role of aerosol transmission of SARS-CoV-2, sampling the air @Liu2020

@@

Viruses are known to transfer from contaminated surfaces to the skin upon contact @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991 and from the skin to the mouth and saliva of individuals @Pitol2017 @Rusin2002. People touch their faces frequently (REF). Therefore, touching contaminated surfaces offers a risk for exposure. Inspite its importance, there is not a lot of information regarding the magnitud of risk associeted with touching contaminated surfaces @BarOn2020.

@@

## Materials and Methods 
### Hazard Identification
When an individual infected with SARS-CoV-2 coughs, sneezes, or speaks, there could be particles of saliva containing the virus that make their way to the proximal surfaces. SARS-CoV-2 can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020. Viruses are known to transfer from surfaces to the skin upon contact  @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991. It is also well known that people frequently touch their faces and the mucouse membranes within the face (nose, eyes and mouth) @Kwok2015. Therefore, indirect transmission of SARS-CoV-2 a relevant transmission pathway. Despite its relevance, there are no studies quantifying the risks associated with indirect transmission of SARS-CoV-2 between surfaces and people. Hence, this risk assessment is designed to explore the risks associated indirect (fomite mediated) transmission of SARS-CoV-2. Since the probability of illness due to infection is, up to date, not known (REF), the end point of this risk assessment is probability of infection.

### Exposure Assessment 
Two expossure scenarios were modeled. 
*Scenario 1. Single touch.*
Frequently touched buttons (Traffic light buttons, train buttons). High touched surfaces and Supermarket products.

@@

#### Parameters table
| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
|:--:|:--:|:--:|:--:|:--:|:--:|
| C~sp~ | Gene Copies/ml | Concentration of SARS-CoV-2 in sputum or saliva | log~10~ [4,4.8...] | ReSample | @Wolfel2020 @Pan2020 @Kim2020, RT-PCR data from 9, 2, and 2 patients with COVID-19|
| GC:TCID~50~ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | @Ip2016, mean RT-PCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
| V~slv/cgh~ | mL | Volume of saliva expelled per cough | |  | |
| N~cgh/min~ |min^-1^ | Number of caughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV| 
|t~i~ | min |Time spend doing the activity *i* | [...] | Normal | This study. Activity ATM: 20 hours of observational data |
|x~i~| m | Inoculation distance for activity *i* | [...] | Point values | Assumed |
| k | PFU^-1^ | Parameter of the exponential dose-response| 0.00246(0.00098) | Normal | http://qmrawiki.org/experiments/sars |
| n | min~-1~ |Decay rate for surface *j*| [...] | Normal | @VanDoremalen2020, CoV-2 infectivity assays, 40% RH and 21-23°C |
| |  ||  | | |

@@

There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals. Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data. Althoug not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF). 
To do this we ...
In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.

#### SARS-CoV-2 Concentration in surfaces

\begin{equation}
C(0)=\frac {V_{slv/cgh} C_{sp}}{4 \pi x_{i}^{2}  0.2}
\end{equation}

\begin{equation}
C(t)= C(0)  e^{-nt}
\end{equation}

#### Dose Response for SARS-CoV
“For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”

The dose-response relationship was obtain from QMRA wiki <http://qmrawiki.org/experiments/sars>. The model is based on animal dose-response relationship. No human dose-response relationship is available for SARS Coronavirus. It is generally accepted to use animal models in QMRA if human data is not available @Haas2014, @Haas2015.

@@

Seasonal factors such as temperature, humidity and sunlight are known to influence the spread of respiratory viruses such as influenza (REF). The role that those factos have on the spread of SARS-CoV-2 remains unknown (REF), althouh it is know that the temperature and humidity influences the viability of the virus (REF), as well as SARS-CoV-2 survive longer at lower temperature (REF).

It has been shown that seasonal environmental factors such as temperature and humidity influence respiratory virus survival and transmission @Moriyama2020.

Humidity influences the size of the droplets generated by the infected individual, the size of the droplets influence both the lifetime of the virus in the environment. 
The deposition of droplets onto surfaces depends on the particle size.

Nevertheless, our results suggest that fomite mediated transmission will be .. influenced by those factors. The extend that this influence will have in the droplet dynanimcs by influencing evaporation rate of the droplets (REF). transmission of CoV will depend on the relative contribution of indirect transmission pathway.

### Limitations

Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020. The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF). In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.

Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.

@@

Currently we have liminted data on the relationship between copy numbers and infectivity, which limits the extend to which the risk assessment can be used. However, modeling approaches such as the QMRA can help elucidate the contribution of different exposure routes to the disease transmission and therefore help the development of guidelines.

## Supporting Information

## Author Information

## Aknowledgements

## References
